# (PART) Nonparametric Estimations{-}

# Parametric vs. Nonparametric methods {-}

According to Breiman, there are two "cultures":  

**The Data Modeling Culture** : One assumes that the data are generated by a given stochastic data model (econometrics) ...  Leo Breiman [@Breiman_TC]: [Statistical Modeling: The Two Cultures](https://projecteuclid.org/download/pdf_1/euclid.ss/1009213726):

> For instance, in the Journal of the American Statistical Association (JASA), virtually every article contains a statement of the form: *Assume that the data are generated by the following model:... I am deeply troubled by the current and past use of data models in applications, where quantitative conclusions are drawn and perhaps policy decisions made.*
>
>... assume the data is generated by independent draws from the model*
>
>$$
>y=b_{0}+\sum_{1}^{M} b_{m} x_{m}+\varepsilon
>$$
>
>where the coefficients are to be estimated. The error term is N(0, $\sigma^2$) and $\sigma^2$ is to be estimated. Given that the data is generated this way, elegant tests of hypotheses,confidence intervals,distributions of the residual sum-of-squares and asymptotics can be derived. This made the model attractive in terms of the mathematics involved. This theory was used both by academics statisticians and others to derive significance levels for coefficients on the basis of model (R), with little consideration as to whether the data on hand could have been generated by a linear model. Hundreds, perhaps thousands of articles were published claiming proof of something or other because the coefficient was significant at the 5% level...
>
>...With the insistence on data models, multivariate analysis tools in statistics are frozen at discriminant analysis and logistic regression in classification and multiple linear regression in regression. Nobody really believes that multivariate data is multivariate normal, but that data model occupies a large number of pages in every graduate text book on multivariate statistical analysis...
>

**Algorithmic Modeling Culture**: One uses algorithmic models and treats the data mechanism as unknown (machine learning) ...    

He argues that the focus in the statistical community on data models has:
  
- Led to irrelevant theory and questionable scientific conclusions; 
- Kept statisticians from using more suitable algorithmic models; 
- Prevented statisticians from working on exciting new problems.  

In **parametric econometrics** we assume that the data come from a generating process that takes the following form:

$$
y=X \beta+\varepsilon
$$

Model ($X$'s) are determined by the researcher and probability theory is a foundation of econometrics


In **Machine learning** we do not make any assumption on how the data have been generated:

$$
y \approx m(X)
$$

Model ($X$'s) is not selected by the researcher and probability theory is not required

Nonparametric econometrics makes the link between the two: **Machine Learning: an extension of nonparametric econometrics**
   
To see the difference between two "cultures", we start with parametric modeling in classification problems.

# Parametric Estimations

So far we have only considered models for numeric response variables. What happens if the response variable is categorical? Can we use linear models in these situations? Yes, we can.  To understand how, let's look at the ordinary least-square (OLS) regression, which is actually a specific case of the more general, generalized linear model (GLM).  So, in general, GLMs relate the mean of the response to a linear combination of the predictors, $\eta(x)$, through the use of a link function, $g(.)$. That is,  

\begin{equation}
\eta(\mathbf{x})=g(\mathrm{E}[Y | \mathbf{X}=\mathbf{x}]),
  (\#eq:5-1)
\end{equation} 

Or,  

\begin{equation}
\eta(\mathbf{x})=\beta_{0}+\beta_{1} x_{1}+\beta_{2} x_{2}+\ldots+\beta_{p-1} x_{p-1} = g(\mathrm{E}[Y | \mathbf{X}=\mathbf{x}])
  (\#eq:5-2)
\end{equation} 
  
In the case of a OLS,  

$$
g(\mathrm{E}[Y | \mathbf{X}=\mathbf{x}]) = E[Y | \mathbf{X}=\mathbf{x}],
$$
  
To illustrate the use of a GLM, we will focus on the case of binary response variable coded using 0 and 1. In practice, these 0s and 1s represent two possible outcomes such as "yes" or "no", "sick" or "healthy", etc.  

$$
Y=\left\{\begin{array}{ll}{1} & {\text { yes }} \\ {0} & {\text { no }}\end{array}\right.
$$

## Linear Probability Models (LPM)

Let's use the dataset **Vehicles** from the `fueleconomy` package.  We we ill create a new variable a new variable, `mpg`, which is 1 if the car's hightway mpg is more than the average, 0 otherwise:  

```{r p1, warning=FALSE, message=FALSE}
library(fueleconomy)  
data(vehicles)
df <- as.data.frame(vehicles)

# Remove NAs
dim(df)
data <- df[complete.cases(df), ]
dim(data)

# Binary outcome mpg = 1 if hwy > mean(hwy), 0 otherwise
data$mpg <- ifelse(data$hwy > mean(data$hwy), 1, 0)
table(data$mpg)
```
  
We are going to have a model that predicts `mpg` (i.e. `mpg` = 1) for each car depending on their attributes. If you check the data, you see that many variables are character variables. Although most functions, like `lm()`, accept character variables (and convert them to factor), it is a good practice to check each variable and convert them appropriate data types.  
  
```{r p2, warning=FALSE, message=FALSE}
for (i in 1:ncol(data)) {
  if(is.character(data[,i])) data[,i] <- as.factor(data[,i])
}
str(data)
```

Done! We are ready to have a model to predict `mpg`.  For now, we'll use only `fuel`.  

```{r p3, warning=FALSE, message=FALSE}
model1 <- lm(mpg ~ fuel + 0, data = data) #No intercept
summary(model1)
```
  
The estimated model is a probabilistic model since,  

$$
E[Y | \mathbf{X}=\mathbf{Regular}]) = \text{Pr}(Y=1|\mathbf{X}=\mathbf{Regular}),
$$
  
In this context, the link function is called "identity" because it directly “links” the probability to the linear function of the predictor variables. Let's see if we can verify this:  

```{r p4, warning=FALSE, message=FALSE}
tab <- table(data$fuel, data$mpg)
ftable(addmargins(tab))
prop.table(tab, 1)
```
  
The frequency table shows the probability of each class (MPG = 1 or 0) for each fuel type.  The OLS we estimated produces exactly the same results, that is,  

$$
Pr[Y = 1 | x=\mathbf{Regular}]) = \beta_{0}+\beta_{1} x_{i}.
$$
  
Since $Y$ has only two possible outcomes (1 and 0), it has a specific probability distribution.  First, let's refresh our memories about Binomial and Bernoulli distributions.  In general, if a random variable, $X$, follows a **binomial distribution** with parameters $n \in \mathbb{N}$ and $p \in [0,1]$, we write $X \sim B(n, p)$. The probability of getting exactly $k$ successes in $n$ trials is given by the probability mass function:  

\begin{equation}
\operatorname{Pr}(X=k)=\left(\begin{array}{l}{n} \\ {k}\end{array}\right) p^{k}(1-p)^{n-k}
  (\#eq:5-3)
\end{equation} 
for $k = 0, 1, 2, ..., n$, where  

$$
\left(\begin{array}{l}{n} \\ {k}\end{array}\right)=\frac{n !}{k !(n-k) !}
$$
  
Formula 5.3 can be understood as follows: $k$ successes occur with probability $p^k$ and $n-k$ failures occur with probability $(1-p)^{n−k}$. However, the $k$ successes can occur anywhere among the $n$ trials, and there are $n!/k!(n!-k!)$ different ways of distributing $k$ successes in a sequence of $n$ trials. Suppose a *biased coin* comes up heads with probability 0.3 when tossed. What is the probability of achieving 4 heads after 6 tosses?  
  
$$
\operatorname{Pr}(4 \text { heads})=f(4)=\operatorname{Pr}(X=4)=\left(\begin{array}{l}{6} \\ {4}\end{array}\right) 0.3^{4}(1-0.3)^{6-4}=0.059535
$$
  
The **Bernoulli distribution** on the other hand, is a discrete probability distribution of a random variable which takes the value 1 with probability $p$ and the value 0 with probability $q = (1 - p)$, that is, the probability distribution of any single experiment that asks a yes–no question.  The **Bernoulli distribution** is a special case of the **binomial distribution**, where $n = 1$. Symbolically, $X \sim B(1, p)$ has the same meaning as $X \sim Bernoulli(p)$. Conversely, any binomial distribution, $B(n, p)$, is the distribution of the sum of $n$ Bernoulli trials, $Bernoulli(p)$, each with the same probability $p$.  

$$
\operatorname{Pr}(X=k) =p^{k}(1-p)^{1-k} \quad \text { for } k \in\{0,1\}
$$
  
Formally, the outcomes $Y_i$ are described as being Bernoulli-distributed data, where each outcome is determined by an unobserved probability $p_i$ that is specific to the outcome at hand, but related to the explanatory variables. This can be expressed in any of the following equivalent forms:  

\begin{equation}
\operatorname{Pr}\left(Y_{i}=y | x_{1, i}, \ldots, x_{m, i}\right)=\left\{\begin{array}{ll}{p_{i}} & {\text { if } y=1} \\ {1-p_{i}} & {\text { if } y=0}\end{array}\right.
  (\#eq:5-4)
\end{equation} 
  
The expression 5.4 is the probability mass function of the Bernoulli distribution, specifying the probability of seeing each of the two possible outcomes. Similarly, this can be written as follows, which avoids having to write separate cases and is more convenient for certain types of calculations. This relies on the fact that $Y_{i}$ can take only the value 0 or 1. In each case, one of the exponents will be 1, which will make the outcome either $p_{i}$ or 1−$p_{i}$, as in 5.4.^[Intuitively, when $n=1$, achieving head once ($k=1$) is $P(head)= p^{k}(1-p)^{1-k}=p$ or $P(tail)= p^{k}(1-p)^{1-k}=1-p.$]  

$$
\operatorname{Pr}\left(Y_{i}=y | x_{1, i}, \ldots, x_{m, i}\right)=p_{i}^{y}\left(1-p_{i}\right)^{(1-y)}
$$
  
Hence this shows that  

$$
\operatorname{Pr}\left(Y_{i}=1 | x_{1, i}, \ldots, x_{m, i}\right)=p_{i}=E[Y_{i}  | \mathbf{X}=\mathbf{x}])
$$

Let's have a more complex model:  

```{r p5, echo=TRUE}
model2 <- lm(mpg ~ fuel + drive + cyl, data = data)
summary(model2)
```
  
Since OLS is a "Gaussian" member of GLS family, we can also estimate it as GLS. We use `glm()` and define the family as "gaussian".  

```{r p6, warning=FALSE, message=FALSE}
model3 <- glm(mpg ~ fuel + drive + cyl, family = gaussian, data = data)
identical(round(coef(model2),2), round(coef(model3),2))
```
  
With this LPM model, we can now predict the classification of future cars in terms of high (`mpg` = 1) or low (`mpg` = 0), which was our objective.  Let's see how successful we are in identifying cars with `mpg` = 1 in our own sample.  

```{r p7, warning=FALSE, message=FALSE}
#How many cars we have with mpg = 1 and mpg = 0 in our data
table(data$mpg) 

#In-sample fitted values or predicted probabilities for mpg = 1
#Remember our E(Y|X) is Pr(Y=1|X)
mpg_hat <- fitted(model2)

#If any predicted mpg above 0.5 should be considered as 1 
length(mpg_hat[mpg_hat > 0.5]) 
length(mpg_hat[mpg_hat <= 0.5])
```
  
Our prediction is significantly off:  we predict many cars with `mpg` = 0 as having `mpg` = 1.

Note that we are using 0.5 as our discriminating threshold to convert predicted probabilities to predicted "labels". This is an arbitrary choice as we will see later

Another issue with LPM can be see below:  

```{r p8, warning=FALSE, message=FALSE}
summary(mpg_hat)
```
  
The predicted probabilities (of `mpg` = 1) are not bounded between 1 and 0.  We will talk about these issues later.  None of these problems are major drawbacks for LPM.  But, by its nature, LPM defines a constant marginal effect of $x$ on $Pr[Y = 1 | x)$.  

$$
Pr(Y = 1 | x=\mathbf{Regular}) = \beta_{0}+\beta_{1} x_{i}.
$$

We can see it with a different example

```{r p9}
model_n <- lm(mpg ~ cyl, data = data)
plot(data$cyl, data$mpg, ylim = c(-1.2, 1.2))
lines(data$cyl, model_n$fitted.values, col = "red", lwd = 2)
```

Two things we see in the plot: first predicted probabilities are not bounded between 0 and 1.  Second, the effect of `cyc` on $Pr(Y = 1 | x)$ is constant regardless of the value of `cyc`.

We can of course add polynomial terms to LPM to deal with it.  But, we may have a better model, Logistic regression, if we think that the constant marginal effect is an unrealistic assumption.

## Logistic Regression

First, let's define some notation that we will use throughout. Note that many machine learning texts use $p$ as the number of parameters.  Here we use it to denote probability.    

$$
p(\mathbf{x})=P(Y=1 | \mathbf{X}=\mathbf{x})
$$

With a binary (Bernoulli) response, we will mostly focus on the case when $Y = 1$, since, with only two possibilities, it is trivial to obtain probabilities when $Y = 0$.  

$$
\begin{array}{c}{P(Y=0 | \mathbf{X}=\mathbf{x})+P(Y=1 | \mathbf{X}=\mathbf{x})=1} \\\\ {P(Y=0 | \mathbf{X}=\mathbf{x})=1-p(\mathbf{x})}\end{array}
$$
  
We begin with introducing the standard logistic function, which is a sigmoid function.  It takes any real input $z$ and outputs a value between zero and one. The standard logistic function is defined as follows:  

\begin{equation}
\sigma(z)=\frac{e^{z}}{e^{z}+1}=\frac{1}{1+e^{-z}}
  (\#eq:5-5)
\end{equation} 

Here is an example:

```{r p10, warning=FALSE, message=FALSE}
set.seed(1)
n <- 500
x = rnorm(n, 0,2)
sigma <- 1/(1+exp(-x))
plot(sigma ~ x, col ="blue", cex.axis = 0.7)
```

This logistic function is nice because: (1) whatever the $x$'s are $\sigma(z)$ is always between 0 and 1, (2) The effect of $x$ on $\sigma(z)$ is not linear.  That is, there is lower and upper thresholds in $x$ that before and after those values (around -2 and 2 here) the marginal effect of $x$ on $\sigma(z)$ is very low.  Therefore, it seems that if we use a logistic function and replace $\sigma(z)$ with $p(x)$, we can solve issues related to these two major drawbacks of LPM.  

Let us assume that $z = y = \beta_{0}+\beta_{1} x_{1}$.  Then, the logistic function can now be written as:  

\begin{equation}
p(x)=P(Y=1|\mathbf{X}=\mathbf{x})=\frac{1}{1+e^{-\left(\beta_{0}+\beta_{1} x\right)}}
  (\#eq:5-6)
\end{equation} 


To understand why nonlinearity would be a desirable future in some probability predictions, let's imagine we try to predict the effect of saving ($x$) on homeownership ($p(x)$).  If you have no saving now ($x=0$), additional \$10K saving would not make a significant difference in your decision to buy a house ($P(Y=1|x)$).  Similarly, when you have \$500K ($x$) saving,  additional \$10K ($dx$) saving should not make a big difference in your decision to buy a house.  That's why flat lower and upper tails of $\sigma(z)$ are nice futures reflecting very low marginal effects of $x$ on the probability of having a house in this case.  

After a simple algebra, we can also write the same function as follows,  

\begin{equation}
\ln \left(\frac{p(x)}{1-p(x)}\right)=\beta_{0}+\beta_{1} x,
  (\#eq:5-7)
\end{equation}

where $p(x)/(1-p(x))$ is called **odds**, a ratio of success over failure.  The natural log of this ratio is called, **log odds**, or **Logit**, usually denoted as $\mathbf(L)$.

```{r p11, warning=FALSE, message=FALSE}
p_x <- sigma
Logit <- log(p_x/(1-p_x)) #By defult log() calculates natural logarithms
plot(Logit ~ x, col ="red", cex.axis = 0.7)
```

In many cases, researchers use a logistic function, when the outcome variable in a regression is dichotomous.  Although there are situations where the linear model is clearly problematic (as described above), there are many common situations where the linear model is just fine, and even has advantages.
  
Let’s start by comparing the two models explicitly. If the outcome $Y$ is dichotomous with values 1 and 0, we define $P(Y=1|X) = E(Y|X)$ as proved earlier, which is just the probability that $Y$ is 1, given some value of the regressors $X$. Then the linear and logistic probability models are:  

$$
P(Y = 1|\mathbf{X}=\mathbf{x})=E(Y | \mathbf{X}=\mathbf{x}) = \beta_{0}+\beta_{1} x_{1}+\beta_{2} x_{2}+\ldots+\beta_{k} x_{k},
$$
$~$

$$
\ln \left(\frac{P(Y=1|\mathbf{X})}{1-P(Y=1|\mathbf{X})}\right)=\beta_{0}+\beta_{1} x_{1}+\ldots+\beta_{k} x_{k}
$$

$~$

While LPM assumes that the probability $P$ is a linear function of the regressors, the logistic model assumes that the natural log of the odds $P/(1-P)$ is a linear function of the regressors.  Note that applying the inverse logit transformation allows us to obtain an expression for $P(x)$.  Finally, LPM can be estimated easily with OLS, the Logistic model needs MLE.  

$$
p(\mathbf{x})=E(Y | \mathbf{X}=\mathbf{x})=P(Y=1 | \mathbf{X}=\mathbf{x})=\frac{1}{1+e^{-(\beta_{0}+\beta_{1} x_{1}+\cdots+\beta_{k} x_{k})}}
$$
$~$

The major advantage of LPM is its interpretability. In the linear model, if $\beta_{2}$ is (say) 0.05, that means that a one-unit increase in $x_{2}$ is associated with a 5-percentage point increase in the probability that $Y$ is 1. Just about everyone has some understanding of what it would mean to increase by 5 percentage points their probability of, say, voting, or dying, or becoming obese.  In the logistic model, however, a change in $x_{1}$ changes the log odds, $\text{log}(P/{(1-P)})$. Hence, the coefficient of a logistic regression requires additional steps to understand what it means: we convert it to the odd ratio (OR) or use the above equation to calculate fitted (predicted) probabilities.

When we should use the logistic model?  It should be the choice if it fits the data much better than the linear model. In other words, **for a logistic model to fit better than a linear model, it must be the case that the log odds are a linear function of X, but the probability is not.**  

Lets review these concepts in a simulation exercise: 

```{r p12, warning=FALSE, message=FALSE}
#Creating random data
set.seed(1)
n <- 500 
x = rnorm(n)
z = -2 + 3 * x

#Probablity is defined by a logistic function
#Therefore it is not a linear function of x!
p = 1 / (1 + exp(-z))

#Remember Bernoulli distribution defines Y as 1 or 0 
y = rbinom(n, size = 1, prob = p)

#And we create our data
data <-  data.frame(y, x)
head(data)
table(y)
```

We know that probability is defined by a logistic function (see above).  What happens if we fit it as LPM, which is $Pr(Y = 1 | x=\mathbf{x}) = \beta_{0}+\beta_{1} x_{i}$?

```{r p13, warning=FALSE, message=FALSE}
lpm <- lm(y ~ x, data = data)
summary(lpm)

plot(x, p, col = "green", 
     cex.lab = 0.7,
     cex.axis = 0.8)
abline(lpm, col = "red")
legend("topleft", c("Estimated Probability by LPM", "Probability"),
       lty = c(1, 1),
       pch = c(NA, NA),
       lwd = 2,
       col = c("red", "green"),
       cex = 0.7)
```

How about a logistic regression?

```{r p14, warning=FALSE, message=FALSE}
logis <- glm(y ~ x, data = data, family = binomial)
summary(logis)

plot(x, p, col = "green", cex.lab = 0.8, cex.axis = 0.8)
curve(predict(logis, data.frame(x), type = "response"),
      add = TRUE,
      col = "red",
      lty = 2)
legend("topleft", c("Estimated Probability by GLM", "Probability"),
       lty = c(1, 1),
       pch = c(NA, NA),
       lwd = 2,
       col = c("red", "green"),
       cex = 0.7)
```

As you can see, the estimated logistic regression coefficients are in line with our DGM coefficients (-2, 3).
  
$$
\log \left(\frac{\hat{p}(\mathbf{x})}{1-\hat{p}(\mathbf{x})}\right)=-1.8253+2.7809 x
$$

### Estimating Logistic Regression

Since Logit is a linear function:  

\begin{equation}
Logit_i = \log \left(\frac{p\left(\mathbf{x}_{\mathbf{i}}\right)}{\left.1-p\left(\mathbf{x}_{\mathbf{i}}\right)\right)}\right)=\beta_{0}+\beta_{1} x_{i 1}+\cdots+\beta_{p-1} x_{i(p-1)},
  (\#eq:5-8)
\end{equation} 

it seems that we can estimate it by a regular OLS.  But, we only observe $Y=1$ or $Y=0$ not $p(\mathbf{x})$. To estimate the $\beta$ parameters, we apply the maximimum likelihood estimation method.  First, we write the likelihood function, $L(\beta)$, given the observed data, which is technically a joint probability density function that can be written a product of $n$ individual density functions:    

$$
L(\boldsymbol{\beta})=\prod_{i=1}^{n} P\left(Y_{i}=y_{i} | \mathbf{X}_{\mathbf{i}}=\mathbf{x}_{\mathbf{i}}\right)
$$
With some rearrangement, we make it more explicit:  

$$
\begin{aligned} L(\boldsymbol{\beta}) &=\prod_{i=1}^{n} p\left(\mathbf{x}_{\mathbf{i}}\right)^{y_{i}}\left(1-p\left(\mathbf{x}_{\mathbf{i}}\right)\right)^{\left(1-y_{i}\right)}  \end{aligned}
$$
With a logarithmic transformation of this function, it becomes a log-likelihood function, which turns products into sums.  Hence, it becomes a linear function:  

\begin{equation} 
\begin{split}
\begin{aligned}
\ell\left(\beta_{0}, \beta\right) &=\sum_{i=1}^{n} y_{i} \log p\left(x_{i}\right)+\left(1-y_{i}\right) \log (1-p\left(x_{i}\right)) \\
&=\sum_{i=1}^{n} \log (1-p\left(x_{i}\right))+\sum_{i=1}^{n} y_{i} \log \frac{p\left(x_{i}\right)}{1-p\left(x_{i}\right)} \\
&=\sum_{i=1}^{n} \log (1-p\left(x_{i}\right))+\sum_{i=1}^{n} y_{i}\left(\beta_{0}+\beta x_{i} \right) \\
&=\sum_{i=1}^{n} \log 1/(1+e^{z_i})+\sum_{i=1}^{n} y_{i}\left(z_i\right) \\
&=\sum_{i=1}^{n} -\log (1+e^{z_i})+\sum_{i=1}^{n} y_{i}\left(z_i\right),
\end{aligned}
\end{split}
(\#eq:5-9)
\end{equation} 
  

where $z_i = \beta_{0}+\beta_{1} x_{1i}+\cdots$.  

Having a function for log-likelihood, we simply need to chose the values of $\beta$ that maximize it. Typically, to find them, we would differentiate the log-likelihood with respect to the parameters ($\beta$), set the derivatives equal to zero, and solve.

$$
\begin{aligned}
\frac{\partial \ell}{\partial \beta_{j}} &=-\sum_{i=1}^{n} \frac{1}{1+e^{\beta_{0}+x_{i}  \beta}} e^{\beta_{0}+x_{i} \beta} x_{i j}+\sum_{i=1}^{n} y_{i} x_{i j} \\
&=\sum_{i=1}^{n}\left(y_{i}-p\left(x_{i} ; \beta_{0}, \beta\right)\right) x_{i j}
\end{aligned}
$$
Unfortunately, there is no closed form for the maximum. However, we can find the best values of $\beta$ by using algorithm (numeric) optimization methods (See Appendix).  

### Cost functions

The cost functions represent optimization objectives in estimations and predictions.  In linear regression, it's a simple sum of squared errors, i.e.  

\begin{equation}
\mathbf{SSE}= \sum{(\hat{y}_i-y_i)^2} 
  (\#eq:5-10)
\end{equation} 
  
If we use a similar cost function in *Logistic Regression* we would have a non-convex function with many local minimum points so that it would be hard to locate the global minimum.  In logistic regression, as we have just seen, the log-likelihood function becomes the cost function. In the machine learning literature notation changes slightly:

\begin{equation}
J =\sum_{i=1}^{n} y_{i} \log p\left(x_{i}\right)+\left(1-y_{i}\right) \log (1-p\left(x_{i}\right)),
  (\#eq:5-11)
\end{equation} 

where for each observation,  

$$
p\left(\mathbf{x}_{\mathbf{i}}\right)=\frac{e^{\beta_{0}+\beta x_{i}}}{1+e^{\beta_{0}+\beta x_{i}}}
$$

Because it is more common to maximize a function in practice, the log likelihood function is inverted by adding a negative sign to the front. For classification problems, equation 5.11 is also called as “log loss“, “cross-entropy” and “negative log-likelihood” used interchangeably.

Now that we have a cost function, we simply need to chose the values of $\beta$ that minimize it. Due to difficulties in multi-dimensional analytic solutions, we use gradient descent and some other types of algorithmic optimization methods.  

The same cost function can be written when $y_i \in \{+1,-1\}$

$$
g_i(\mathbf{w})= \begin{cases}-\log \left(p\left({\mathbf{x}_i}^{T} \mathbf{w}\right)\right) & \text { if } y_{i}=+1 \\ -\log \left(1-p\left({\mathbf{x}_i}^{T} \mathbf{w}\right)\right) & \text { if } y_{i}=-1\end{cases}
$$
  
We can then form the *Softmax* cost for Logistic regression by taking an average of these Log Error costs as
  
$$
g(\mathbf{w})=\frac{1}{n} \sum_{i=1}^{n} g_{i}(\mathbf{w}) .
$$
  
It is common to express the Softmax cost differently by re-writing the Log Error in a equivalent way as follows. Notice that with $z = \mathbf{x}^{T} \mathbf{w}$
    
$$
1-p(z)=1-\frac{1}{1+e^{-z}}=\frac{1+e^{-z}}{1+e^{-z}}-\frac{1}{1+e^{-z}}=\frac{e^{-z}}{1+e^{-z}}=\frac{1}{1+e^{z}}=p(-z)
$$
  
Hence, the point-wise cost function can be written as

$$
g_{i}(\mathbf{w})= \begin{cases}-\log \left(p\left({\mathbf{x}}_{i}^{T} \mathbf{w}\right)\right) & \text { if } y_{i}=+1 \\ -\log \left(p\left(-{\mathbf{x}}_{i}^{T} \mathbf{w}\right)\right) & \text { if } y_{i}=-1\end{cases}
$$

Now notice that because we are using the *label* values $\pm 1$ we can move the label value in each case inside the inner most parenthesis, 

$$
g_{i}(\mathbf{w})=-\log \left(p\left(y_{i} {\mathbf{x}}_{i}^{T} \mathbf{w}\right)\right)
$$
Finally since $-\log (x)=\frac{1}{x}$, we can re-write the point-wise cost above equivalently as
  
$$
g_{i}(\mathbf{w})=\log \left(1+e^{-y_{i} \mathbf{x}_{i}^{T} \mathbf{w}}\right)
$$
  
The average of this point-wise cost over all $n$ points we have the common Softmax cost for logistic regression: 
  
$$
g(\mathbf{w})=\frac{1}{n} \sum_{i=1}^{n} g_{i}(\mathbf{w})=\frac{1}{n} \sum_{i=1}^{n} \log \left(1+e^{-y_{i} \mathbf{x}_{i}^{T} \mathbf{w}}\right)
$$
  
This will be helpful when we make the comparisons between logistic regression and support vector machines in Chapter 15.  
  
### Deviance
You have probably noticed that the output from `summary()` reports the “deviance” measures for logistic regressions. The "Null deviance" is the deviance for the null model, that is, a model with no predictors. The null deviance shows how well the response variable is predicted by a model that includes only the intercept (grand mean).  What is **deviance**? 

It is defined as the difference of likelihoods between the fitted model and the saturated model:  

\begin{equation}
D=-2 \ell(\hat{\beta})+2 \ell(\text { saturated model })
  (\#eq:5-12)
\end{equation} 
  
This is also known as the *Likelihood Ratio Test* (LRT) that has been used to compare two nested models.  

\begin{equation}
\mathbf{L R T}=-2 \log\left(\frac{L_{s}(\hat{\theta})}{L_{g}(\hat{\theta})}\right)
  (\#eq:5-13)
\end{equation} 
  
where $L_s$ in 5.13 is the likelihood for the null model and $L_g$ is the likelihood for the alternative model.  

The perfect model, known as the saturated model, denotes an abstract model that fits perfectly the sample, that is, the model such that $P(Y=1 | \mathbf{X}=\mathbf{x})=Y_{i}$.  As the likelihood of the saturated model is exactly one, the deviance can be expressed as  

$$
D=-2 \ell(\hat{\beta})
$$
Therefore, the deviance is always larger than or equal to zero, which means a perfect fit. We can evaluate the magnitude of the deviance relative to the null deviance,  

$$
D_{0}=-2 \ell\left(\hat{\beta}_{0}\right),
$$

reflecting the deviance of the worst model, which has no predictors. Hence, this comparison shows how much our fitted model has improved relative to the benchmark. We can develop a metric, the (Pseudo) $R^{2}$ statistic:  

\begin{equation}
R^{2}=1-\frac{D}{D_{0}}
  (\#eq:5-14)
\end{equation} 
  
Similar to $R^{2}$,  the (Pseudo) $R^{2}$ is a quantity between 0 and 1. If the fit is perfect, then $D = 0$ and $R^{2}=1$.  

### Predictive accuracy

Another way of evaluating the model's fit is to look at its predictive accuracy. When we are interested simply in prediction in classification, but not in predicting the value of $\hat{p(x)}$, such as  

$$
\hat{Y}=\left\{\begin{array}{ll}{1,} & {\hat{p}\left(x_{1}, \ldots, x_{k}\right)>\frac{1}{2}} \\ {0,} & {\hat{p}\left(x_{1}, \ldots, x_{k}\right)<\frac{1}{2}}\end{array}\right.
$$
then, the overall predictive accuracy can be summarized with a matrix,  

$$
\begin{array}{ccc}{\text { Predicted vs. Reality}} & {Y=1} & {Y=0} \\ {\hat{Y}=1} & {\text { TP }_{}} & {\text { FP }_{}} \\ {\hat{Y}=0} & {\text { FN }_{}} & {\text { TN }_{}}\end{array}
$$

where, TP, FP, FN, TN are **"True positives", "False Positives", "False Negatives", "True Negatives"**, respectively.  This table is also know as **Confusion Table**.  There are many metrics that can be calculated from this table to measure the accuracy of our classifier.  We will spend more time on this this subject under Chapter 10 later.

