# Lasso

The penalty in ridge regression, $\lambda \sum_{j} \beta_{j}^{2}$, will shrink all of the coefficients towards zero, but it will not set any of them exactly to zero.  This may present a problem in model interpretation when the number of variables is quite large.  One of the key advantages of Lasso is that it can set the coefficients of some features to exactly zero, effectively eliminating those features from the model.
  
By eliminating unnecessary or redundant features from the model, Lasso can help to improve the interpretability and simplicity of the model. This can be particularly useful when you have a large number of features and you want to identify the most important ones for predicting the target variable.

The lasso, a relatively recent alternative to ridge regression, minimizes the following quantity:

\begin{equation}
\sum_{i=1}^{n}\left(y_{i}-\beta_{0}-\sum_{j=1}^{p} \beta_{j} x_{i j}\right)^{2}+\lambda \sum_{j=1}^{p}\left|\beta_{j}\right|=\operatorname{RSS}+\lambda \sum_{j=1}^{p}\left|\beta_{j}\right|
  (\#eq:16-1)
\end{equation} 

The lasso also shrinks the coefficient estimates towards zero. However, the $\ell_{1}$ penalty, the second term of equation 18.1, has the effect of forcing some of the coefficient estimates to be exactly equal to zero when the tuning parameter $\lambda$ is sufficiently large.  Hence, the lasso performs variable selection. As a result, models generated from the lasso are generally much easier to interpret than those produced by ridge regression.  

In general, one might expect lasso to perform better in a setting where a relatively small number of predictors have substantial coefficients and the remaining predictors have no significant effect on the outcome. This property is known as "sparsity", because it results in a model with a relatively small number of non-zero coefficients.  In some cases, Lasso can find a true sparsity pattern in the data by identifying a small subset of the most important features that are sufficient to accurately predict the target variable.

Now, we apply lasso to the same data, `Hitters`.  Again, we will follow a similar way to compare ridge and lasso as in [Introduction to Statistical Learning](https://www.statlearning.com) (ISLR).    

```{r, warning=FALSE, message=FALSE, cache=TRUE}
library(glmnet)
library(ISLR)
remove(list = ls())

data(Hitters)
df <- Hitters[complete.cases(Hitters$Salary), ]
X  <- model.matrix(Salary ~ ., df)[,-1]
y <- df$Salary

# Without a specific grid on lambda
set.seed(1)
train <- sample(1:nrow(X), nrow(X) * 0.5)
test <- c(-train)
ytest <- y[test]

# Ridge
set.seed(1)
ridge.out <- cv.glmnet(X[train,], y[train], alpha = 0)
yhatR <- predict(ridge.out, s = "lambda.min", newx = X[test,])
mse_r <- mean((yhatR - ytest)^2)

# Lasso
set.seed(1)
lasso.out <- cv.glmnet(X[train,], y[train], alpha = 1)
yhatL <- predict(lasso.out, s = "lambda.min", newx = X[test,])
mse_l <- mean((yhatL - ytest) ^ 2)

mse_r
mse_l
```

Now, we will define our own grid search:  

```{r, warning=FALSE, message=FALSE, cache=TRUE}
# With a specific grid on lambda + lm()
grid = 10 ^ seq(10, -2, length = 100)

set.seed(1)
train <- sample(1:nrow(X), nrow(X)*0.5)
test <- c(-train)
ytest <- y[test]

#Ridge
ridge.mod <- glmnet(X[train,], y[train], alpha = 0,
                    lambda = grid, thresh = 1e-12)
set.seed(1)
cv.outR <- cv.glmnet(X[train,], y[train], alpha = 0)
bestlamR <- cv.outR$lambda.min
yhatR <- predict(ridge.mod, s = bestlamR, newx = X[test,])
mse_R <- mean((yhatR - ytest) ^ 2)

# Lasso
lasso.mod <- glmnet(X[train,], y[train], alpha = 1,
                    lambda = grid, thresh = 1e-12)
set.seed(1)
cv.outL <- cv.glmnet(X[train,], y[train], alpha = 1)
bestlamL <- cv.outL$lambda.min
yhatL <- predict(lasso.mod, s = bestlamL, newx = X[test,])
mse_L <- mean((yhatL - ytest) ^ 2)

mse_R
mse_L
```

Now, we apply our own algorithm:  

```{r, warning=FALSE, message=FALSE, cache=TRUE}
grid = 10 ^ seq(10, -2, length = 100)
MSPE <- c()
MMSPE <- c()

for(i in 1:length(grid)){
  for(j in 1:100){
    set.seed(j)
    ind <- unique(sample(nrow(df), nrow(df), replace = TRUE))
    
    train <- df[ind, ]
    xtrain <- model.matrix(Salary ~ ., train)[,-1]
    ytrain <- df[ind, 19]
    
    test <- df[-ind, ]
    xtest <- model.matrix(Salary~., test)[,-1]
    ytest <- df[-ind, 19]
  
    model <- glmnet(xtrain, ytrain, alpha = 1,
                    lambda = grid[i], thresh = 1e-12)
    yhat <- predict(model, s = grid[i], newx = xtest)
    MSPE[j] <- mean((yhat - ytest) ^ 2)
    }
  MMSPE[i] <- mean(MSPE)
}

min(MMSPE)
grid[which.min(MMSPE)]
plot(log(grid), MMSPE, type="o", col = "red", lwd = 3)
```

What are the coefficients?

```{r, warning=FALSE, message=FALSE}
coef_lasso <- coef(model, s=grid[which.min(MMSPE)], nonzero = T)
coef_lasso
```

We can also try a classification problem with LPM or Logistic regression when the response is categorical. If there are two possible outcomes, we use the binomial distribution, else we use the multinomial.

# Adaptive Lasso

Adaptive lasso is a method for regularization and variable selection in regression analysis that was introduced by Zou (2006) in [The Adaptive Lasso and Its Oracle Properties](http://users.stat.umn.edu/~zouxx019/Papers/adalasso.pdf). In this paper, the author proposed the use of a weighted $\ell_{1}$ penalty in the objective function, with the weights chosen to adapt to the correlation structure of the data. He showed that this method can result in a more stable model with fewer coefficients being exactly zero, compared to the standard lasso method which uses a simple $\ell_{1}$ penalty. 

Since its introduction, adaptive lasso has been widely used in a variety of applications in statistical modeling and machine learning. It has been applied to problems such as feature selections in genomic data, high-dimensional regressions, and model selections with generalized linear models.  Adaptive lasso is useful in situations where the predictors are correlated and there is a need to select a small subset of important variables to include in the model. It has been shown that adaptive lasso is an oracle efficient estimator (consistency in variable selection and asymptotic normality in coefficient estimation), while the plain lasso is not.
  
Consider the linear regression model:

$$
y_i=x_i^{\prime} \beta+\epsilon_i, ~~~~i=1, \ldots, n ~~~~\text{and} ~~~~\beta \text { is } (p \times 1)
$$
The adaptive Lasso estimates $\beta$ by minimizing
  
$$
L(\beta)=\sum_{i=1}^n\left(y_i-x_i^{\prime} \beta\right)^2+\lambda_n \sum_{j=1}^p \frac{1}{w_j}\left|\beta_j\right|
$$
  
where, typically $w_j=(\left|\hat{\beta}_{O L S_j}\right|)^{\gamma}$ or $w_j=(\left|\hat{\beta}_{Ridge_j}\right|)^{\gamma}$, where $\gamma$ is a positive constant for adjustment of the Adaptive Weights vector, and suggested to be the possible values of 0.5, 1, and 2.

The weights in adaptive lasso (AL) provides a prior "intelligence" about variables such that,while the plain Lasso penalizes all parameters equally, the adaptive Lasso is likely to penalize non-zero coefficients less than the zero ones. This is because the weights can be obtained from the consistent least squares estimator. If $\beta_{AL, j}=0$, then $\hat{\beta}_{O L S, j}$ is likely to be close to zero leading to a very small $w_j$. Hence, truly zero coefficients are penalized a lot.  Calculating the weights in adaptive lasso requires a two-step procedure

Here is an example where we use the ridge weight in adaptive lasso:  

```{r, message=FALSE, warning=FALSE}
library(ISLR)
library(glmnet)

remove(list = ls())

data(Hitters)
df <- Hitters[complete.cases(Hitters$Salary), ]
X  <- model.matrix(Salary~., df)[,-1]
y <- df$Salary

# Ridge weights with gamma = 1
g = 1
set.seed(1)
modelr <- cv.glmnet(X, y, alpha = 0)
coefr <- as.matrix(coef(modelr, s = modelr$lambda.min))
w.r <- 1/(abs(coefr[-1,]))^g

## Adaptive Lasso
set.seed(1)
alasso <- cv.glmnet(X, y, alpha=1, penalty.factor = w.r)

## Lasso
set.seed(1)
lasso <- cv.glmnet(X, y, alpha=1)

# Sparsity
cbind(LASSO = coef(lasso, s="lambda.1se"),
           ALASSO = coef(alasso, s="lambda.1se"))
```

We can see the difference between lasso and adaptive lasso in this example: `PutOuts`, `CRuns`, and `CRBI` picked by lasso are not selected by adaptive lasso. There are only three common features in both methods: `Hits`, `Walks`, and `DivisionW`. To understand which model is better in terms of catching the true sparsity, we will have a simulation to illustrate some of the properties of the Lasso and the adaptive Lasso.

# Sparsity

This is a simulation to illustrate some of the properties of Lasso-type estimations. There are two objectives in using these penalized regressions: model selection (identifying "correct" sparsity) and prediction accuracy.  These two objectives require different optimization approaches and usually are not compatible. In model selection, the objective is to shrink the dimension of the model to the “true” sparsity. This is usually evaluated by checking whether the Oracle properties are satisfied.  These asymptotic properties look at (1) if the model identified by the penalized regression converges to the “true” sparsity, (2) if the coefficients are consistent.

The literature suggests that Lasso is not an oracle estimator. Adaptive Lasso was developed (Zou 2006) to fill this gap.

Let's specify a data generating process with a linear regression model:

$$
y_i=x_i^{\prime} \beta+u_i, ~~~~~i=1, \ldots, n
$$

where $\beta$ is $p \times 1$. First, we consider the case where $p<n$ then move to the case where $p \geq n$.  We define $\beta=(1,1,0,0)^{\prime}$ and $n=100$.

```{r}
#This function generates the data
dgp <- function(N, Beta) {
  p = length(Beta)
  
  X <- matrix(rnorm(N * p), ncol = p)
  u <- matrix(rnorm(N), ncol = 1)
  dgm <- X %*% Beta
  y <- X %*% Beta + u
  
  return <- list(y, X)
}

N = 100
Beta = c(1, 1, 0, 0)

set.seed(148)
Output <- dgp(N, Beta)
y <- Output[[1]]
X <- Output[[2]]
```

First, we apply lasso

```{r, message=FALSE, warning=FALSE}
library(glmnet)

set.seed(432)
lasso <- glmnet(x = X, y = y, family = "gaussian")

beta_hat <- lasso$beta
S_matrix <- cbind(t(beta_hat), "lambda" = lasso$lambda)
S_matrix[c(1:8, 25:30, 55:60), ] # selected rows
```
  
Which set of beta_hat should we select?  To answer this question we need to find the lambda.  We need $\lambda_n \rightarrow \infty$ in order to shrink the truly zero coefficients to zero. This requires $\lambda_n$ to be sufficiently large. This would introduce asymptotic bias to the non-zero coefficients. 
  
In practice, choosing $\lambda_n$ by $\mathrm{BIC}$ (Bayesian Information Criterion) results in a consistent model selection in the fixed $p$ setting. That is, let $\mathcal{A}=\left\{j: \beta_{0, j} \neq 0\right\}$,  active set or relevant variables,
  
$$
P\left(\hat{\mathcal{A}}_{\lambda_{BIC}}=\mathcal{A}\right) \rightarrow 1
$$

Thus, let $S S E_\lambda$ be the sum of squared error terms for a given value of $\lambda$ and $n z_\lambda$ be the number of non-zero coefficients. Then, it can be shown that
  
$$
B I C_\lambda=\log \left(S S E_\lambda\right)+\frac{\log (n)}{n} n z_\lambda
$$

```{r, message=FALSE, warning=FALSE}
# Predict yhat for each of 61 lambda (s)
y_hat = predict(lasso, newx = X)
dim(y_hat)

# SSE for each lambda (s)
SSE <- c()
for (i in 1:ncol(y_hat)) {
  SSE_each <- sum((y_hat[, i] - y[, 1]) ^ (2))
  SSE <- c(SSE, SSE_each)
}

# BIC
nz <- colSums(beta_hat != 0) # Number of non-zero coefficients for each lambda
BIC <- log(SSE) + (log(N) / N) * nz # BIC
BIC
```
  
And, the selected model that has the minimum BIC

```{r, message=FALSE, warning=FALSE}
beta_lasso <- beta_hat[, which(BIC == min(BIC))]
beta_lasso
```

This is the `beta_hat` that identifies the true sparsity. And, the second Oracle property, the $\ell_2$ error:

```{r, message=FALSE, warning=FALSE}
l_2 <- sqrt(sum((beta_lasso - Beta) ^ 2))
l_2
```

Here we will create a simulation that will report two Oracle Properties for Lasso and Adaptive Lasso:

- True sparsity,  
- $\ell_2$  error.  

**Lasso**
  
We first have a function, `msc()`, that executes a simulation with all the steps shown before:

```{r, message=FALSE, warning=FALSE}
mcs <- function(mc, N, Beta) {
  mcmat <- matrix(0, nrow = mc, ncol = 3)
  beta_lasso_mat <- matrix(0, nr = mc, nc = length(Beta))
  
  for (i in 1:mc) {
    set.seed(i)
    data <- dgp(N, Beta)
    y <- data[[1]]
    X <- data[[2]]
    
    set.seed(i)
    lasso <- glmnet(x = X, y = y, family = "gaussian")
    beta_hat <- lasso$beta    # beta_hat is a matrix
    y_hat = predict(lasso, newx = X)
    
    SSE <- c()
    for (j in 1:ncol(y_hat)) {
      SSE_each <- sum((y_hat[, j] - y[, 1]) ^ (2))
      SSE <- c(SSE, SSE_each)
    }
    
    nz <- colSums(beta_hat != 0)
    BIC <- log(SSE) + (log(N) / N) * nz
    beta_lasso <- beta_hat[, which(BIC == min(BIC))]
    nonz_beta = length(Beta[Beta == 0])
    nonz_beta_hat = length(beta_lasso[beta_lasso == 0])
    
    mcmat[i, 1] <- sqrt(sum((beta_lasso - Beta) ^ 2))
    mcmat[i, 2] <- ifelse(nonz_beta != nonz_beta_hat, 0, 1)
    mcmat[i, 3] <- sum(beta_lasso != 0)
    beta_lasso_mat[i, ] <- beta_lasso
  }
  return(list(mcmat, beta_lasso_mat))
}
```

We are ready for simulation:

```{r, message=FALSE, warning=FALSE}
mc <- 500
N <- 1000
Beta <- matrix(c(1, 1, 0, 0), nc = 1)
output <- mcs(mc, N, Beta) #see the function

MC_betas = output[[2]]
MC_performance = output[[1]]

sum(MC_performance[, 2]) #how many times lasso finds true sparsity
```

This is the first property: lasso identifies the true sparsity $400/500 = 80\%$ of cases. And the second property, $\ell_2$ error, in the simulation is (in total):  

```{r}
sum(MC_performance[, 1])
```

**Adaptive Lasso**
  
This time we let our adaptive lasso use lasso coefficients as penalty weights in `glmnet()`.   Let's have the same function with Adaptive Lasso for the simulation:

```{r}
# Adaptive LASSO
mcsA <- function(mc, N, Beta) {
  mcmat <- matrix(0, nr = mc, nc = 3)
  beta_lasso_mat <- matrix(0, nr = mc, nc = length(Beta))
  
  for (i in 1:mc) {
    data <- dgp(N, Beta)
    y <- data[[1]]
    X <- data[[2]]
    
    lasso <- glmnet(x = X, y = y, family = "gaussian")
    beta_hat <- lasso$beta
    
    y_hat = predict(lasso, newx = X)
    
    SSE <- c()
    for (j in 1:ncol(y_hat)) {
      SSE_each <- sum((y_hat[, j] - y[, 1]) ^ (2))
      SSE <- c(SSE, SSE_each)
    }
    
    nz <- colSums(beta_hat != 0)
    BIC <- log(SSE) + (log(N) / N) * nz
    beta_lasso <- beta_hat[, which(BIC == min(BIC))]
    
    weights = abs(beta_lasso) ^ (-1)
    weights[beta_lasso == 0] = 10 ^ 10 # to handle inf's
    
    #Now Adaptive Lasso
    lasso <-
      glmnet(
        x = X,
        y = y,
        family = "gaussian",
        penalty.factor = weights
      )
    beta_hat <- lasso$beta
    
    y_hat = predict(lasso, newx = X)
    
    SSE <- c()
    for (j in 1:ncol(y_hat)) {
      SSE_each <- sum((y_hat[, j] - y[, 1]) ^ (2))
      SSE <- c(SSE, SSE_each)
    }
    
    nz <- colSums(beta_hat != 0)
    BIC <- log(SSE) + (log(N) / N) * nz
    beta_lasso <- beta_hat[, which(BIC == min(BIC))]
    nonz_beta = length(Beta[Beta == 0])
    nonz_beta_hat = length(beta_lasso[beta_lasso == 0])
    
    mcmat[i, 1] <- sqrt(sum((beta_lasso - Beta) ^ 2))
    mcmat[i, 2] <- ifelse(nonz_beta != nonz_beta_hat, 0, 1)
    mcmat[i, 3] <- sum(beta_lasso != 0)
    beta_lasso_mat[i, ] <- beta_lasso
  }
  return(list(mcmat, beta_lasso_mat))
}
```

Here are the results for adaptive lasso:  

```{r}
mc <- 500
N <- 1000
beta <- matrix(c(1, 1, 0, 0), nc = 1)
output <- mcsA(mc, N, beta) #see the function

MC_betas = output[[2]]
MC_performance = output[[1]]

sum(MC_performance[, 2])
```

And,

```{r}
sum(MC_performance[,1])
```

The simulation results clearly show that Adaptive Lasso is an Oracle estimator and a better choice for sparsity applications.
  
We saw here a basic application of adaptive lasso, which has several different variations in practice, such as Thresholded Lasso and Rigorous Lasso.  Model selections with lasso has been an active research area.  One of the well-known applications is the double-selection lasso linear regression method that can be used for variable selections.  Moreover, lasso type applications are also used in time-series forecasting and graphical network analysis for dimension reductions. 


