# Principle Component Analysis

Having seen SVD and Eigenvalue decomposition, now we can look at Principle Component Analysis (PCA), which is a statistical procedure that allows you to summarize the information content in large data tables.  In other words, it helps dimension reduction in big datasets.

**PCA is a eigenvalue decomposition of a covariance matrix** (of data matrix $\mathbf{X}$). Since a covariance matrix is a square matrix, we can apply the eigenvalue decomposition, which reveals the unique orthogonal directions (variances) in the data so that their orthogonal linear combinations maximize the total variance.

The goal is here a dimension reduction of the data matrix.  Hence by selecting a few loading, we can reduce the dimension of the data but capture a substantial variation in the data at the same time.  See <https://www.youtube.com/watch?v=fkf4IBRSeEc> [@Brunton_PCA] and 
<https://setosa.io/ev/principal-component-analysis/> [@Powell_PCA]. 

Principal components are the ordered (orthogonal) lines (vectors) that best account for the maximum variance in the data by their magnitude. To get the (unique) variances (direction and the magnitude) in data, we first obtain the mean-centered covariance matrix.  And as you can imagine, eigenvectors (which give the unique directions) and eigenvalues (which identify those directions' magnitude) are used for PCA:

So when we use the covariance matrix of a data, we can use eigenvalue decomposition to identify the unique variation and their relative magnitude in the data.  Here is a simple procedure:  
  
1. $\mathbf{X}$ is the data matrix,  
2. $\mathbf{B}$ is the mean-centered data matrix,  
3. $\mathbf{C}$ is the covariance matrix ($\mathbf{B^TB}$) (note if $\mathbf{B}$ is scaled, i.e. "z-scored", $\mathbf{B^TB}$ gives correlation matrix)
4. Compute the eigenvectors and values of $\mathbf{C}$: $\mathbf{C} = \mathbf{VDV^{\top}}$ hence $\mathbf{CV} = \mathbf{VD}$, where $\mathbf{V}$ is the eigenvectors (loadings) and $\mathbf{D}$ is eigenvalues.
5. Using $\mathbf{V}$, the transformation of $\mathbf{B}$ with $\mathbf{B} \mathbf{V}$ maps the data of $p$ variables to a new space of $p$ variables which are uncorrelated over the dataset. $\mathbf{T} (=\mathbf{B} \mathbf{V})$ is called the **principle component or score matrix**
6. Since SVD of $\mathbf{B} = \mathbf{U} \Sigma \mathbf{V}^{\top}$, we can also get $\mathbf{B}\mathbf{V} = \mathbf{T} = \mathbf{U\Sigma}$. Hence the principle components are $\mathbf{T} = \mathbf{BV} = \mathbf{U\Sigma}$.
7. However, not all the principal components need to be kept. Keeping only the first $r$ principal components, produced by using only the first $r$ eigenvectors, gives the truncated transformation $\mathbf{T}_{r} = \mathbf{B} \mathbf{V}_{r}$.  Obviously you choose those with higher variance in each directions by the order of eigenvalues.
8. We can use $\frac{\lambda_{k}}{\sum_{i=1} \lambda_{k}}$ to identify $r$. Or cumulatively, we can see how much variation could be captured by $r$ number of $\lambda$s, which gives us an idea how many principle components to keep ...

$$
\frac{\sum_{i=1}^{r} \lambda_{k}}{\sum_{i=1}^n \lambda_{k}}
$$

Let's see an example.  Here is the data:

```{r pca1, warning=FALSE, message=FALSE}
library("factoextra")
data(decathlon2)

X <- as.matrix(decathlon2[, 1:10])
head(X)

n <- nrow(X)

B <- scale(X, center = TRUE)
C <- t(B)%*%B/(n-1)
head(C)
#Check it
head(cov(B))
```

Eigenvalues and vectors ...

```{r, warning=FALSE, message=FALSE}
#Eigens
evalues <- eigen(C)$values
evalues
evectors <- eigen(C)$vectors
evectors #Ordered
```

Now with `prcomp()`.  First, eigenvalues:  

```{r pca2, warning=FALSE, message=FALSE}
# With `prcomp()`
Xpca <- prcomp(X, scale = TRUE)
#Eigenvalues
Xpca$sdev # you can see it's ordered
# They are sqrt() of eigenvalues that we calculated earlier
sqrt(evalues)
```

Loadings ...

```{r pca3, warning=FALSE, message=FALSE}
#Eigenvectors (loadings)
Xpca$rotation # 10x10
loadings <- Xpca$rotation
```

Interestingly the signs of eigenvectors are flipped and opposites of what we calculated with `eigen()` above.  There are multiple discussions about the sign reversals in eignevectores.  You can find them [here](https://stats.stackexchange.com/questions/154716/pca-eigenvectors-of-opposite-sign-and-not-being-able-to-compute-eigenvectors-wi) [@Kroll_2015] and [here](https://stackoverflow.com/questions/55076133/dont-know-why-eigen-gives-a-vectors-of-wrong-sign-and-the-loading-matrix-is-j) [@Wilks_2019]

Visualize the order ... 

```{r pca4, warning=FALSE, message=FALSE}
plot(Xpca$sdev) # Eigenvalues
fviz_eig(Xpca) # Cumulative with "factoextra"

# Or
var <- (Xpca$sdev)^2
var_perc <- var/sum(var) * 100

barplot(var_perc, xlab='PC', ylab='Percent Variance',
        names.arg=1:length(var_perc), las=1,
        ylim=c(0, max(var_perc)), col='lightgreen')
 
abline(h=mean(var_perc), col='red')
```
  
Since we have ten variables, if each variable contributed equally, they would each contribute 10\% to the total variance (red line). This criterion suggests we should also include principal component 4 (but barely) in our interpretation. 

And principle component scores $\mathbf{T} = \mathbf{X}\mathbf{V}$ (a.k.a score matrix) with `prcomp()`:

```{r pca5, warning=FALSE, message=FALSE}
pc <- scale(X)%*%Xpca$rotation
head(pc)
dim(pc)

# which is also given by `prcomp()`
head(Xpca$x)
```

Now you can think that if we use `evectors` that we calculated earlier with filliped signs, the data would be different.  It's similar to multiply the entire data with -1.  So the data would not change in a sense that that captures the variation between observations and variables.  That's why the sign of eigenvalues are arbitraray.

With SVD ...
  
```{r pca6, warning=FALSE, message=FALSE}
# With SVD
Xsvd <- svd(scale(X))
pc_2 <- Xsvd$u%*%diag(Xsvd$d)
dim(pc_2)
head(pc_2)
```

Here we can reduce the dimensionality by selecting only 4 PC (the first 4 PC's are above the average, which explain more than 80% of the variation in the data - see the graph above)

```{r pca7, warning=FALSE, message=FALSE}
reduced <- pc[, 1:4]
dim(reduced)
head(reduced)
```

The individual columns of $\mathbf{T}$ successively inherit the maximum possible variance from $\mathbf{X}$, with each coefficient vector in $\mathbf{V}$ constrained to be a unit vector. The full principal components decomposition of $\mathbf{X}$, $\mathbf{T}=\mathbf{X V}$, where $\mathbf{V}$ is a $p \times p$ matrix of weights whose columns are the eigenvectors of $\mathbf{X}^{\top} \mathbf{X}$. Columns of $\mathbf{V}$ multiplied by the square root of corresponding eigenvalues, **that is, eigenvectors scaled up by the variances, are called loadings in PCA or in Factor analysis**.

Note that if we make a singular value decomposition for a covariance matrix 

$$
\begin{aligned}
\mathbf{X}^{T} \mathbf{X} &=\mathbf{V} \mathbf{\Sigma}^{\top} \mathbf{U}^{\top} \mathbf{U} \mathbf{\Sigma} \mathbf{V}^{\top} \\
&=\mathbf{V} \mathbf{\Sigma}^{\top} \mathbf{\Sigma} \mathbf{V}^{\top} \\
&=\mathbf{V} \hat{\mathbf{\Sigma}}^{2} \mathbf{V}^{\top}
\end{aligned}
$$

where $\hat{\boldsymbol{\Sigma}}$ is the square diagonal matrix with the singular values of $\mathbf{X}$ and the excess zeros chopped off that satisfies $\hat{\boldsymbol{\Sigma}}^{2}=\boldsymbol{\Sigma}^{\top} \boldsymbol{\Sigma}$.
  
Comparison with the eigenvector factorization of $\mathbf{X}^{\top} \mathbf{X}$ establishes that the right singular vectors $\mathbf{V}$ of $\mathbf{X}$ are equivalent to the eigenvectors of $\mathbf{X}^{\top} \mathbf{X}$, while the singular values $\sigma_{(k)}$ of $\mathbf{X}$ are equal to the square-root of the eigenvalues $\lambda_{(k)}$ of $\mathbf{X}^{\top} \mathbf{X}$.

```{r pca8}
biplot(reduced[, 1:2], loadings[, 1:2], cex=0.7)
```
