# (PART) Network Analysis {-}

# Graphical Network Analysis {-}
  
Following chapters are work in progress without a proper copyediting ...

# Fundementals

Graphical modeling refers to a class of probabilistic models that uses graphs to express conditional (in)dependence relations between random variables. A graphical model represents the probabilistic relationships among a set of variables. Nodes in the graph correspond to variables, and the absence of edges (partial correlations) corresponds to conditional independence. Graphical models are becoming more popular in statistics and in its applications in many different fields for several reasons.

The central idea is that each variable is represented by a node in a graph. Any pair of nodes may be joined by an edge. For most types of graph a missing edge represents some form of independency between the pair of variables. Because the independency may be either marginal or conditional on some or all of the other variables, a variety of types of graph are needed.

A particularly important distinction is between directed and undirected edges. In the former an arrow indicates the direction of dependency from an explanatory variable to a response. If, however, the two variables are to be interpreted on an equal footing then an edge between them is undirected, or cyclic dependencies are permitted. 

Hence, we need to cover several concept related to statistical dependece, correlations

## Covariance

```{r gna1, warning=FALSE, message=FALSE}
library(ppcor)
library(glasso)
library(glassoFast)
library(corpcor)
library(rags2ridges)
```

First, the data matrix, which refers to the array of numbers:  

$$
\mathbf{X}=\left(\begin{array}{cccc}
x_{11} & x_{12} & \cdots & x_{1 p} \\
x_{21} & x_{22} & \cdots & x_{2 p} \\
x_{31} & x_{32} & \cdots & x_{3 p} \\
\vdots & \vdots & \ddots & \vdots \\
x_{n 1} & x_{n 2} & \cdots & x_{n p}
\end{array}\right)
$$

Our data

```{r gna2}
set.seed(5)
x <- rnorm(30, sd=runif(30, 2, 50))
X <- matrix(x, 10)
X
```

The Covariation of Data

$$
\mathbf{S}=\left(\begin{array}{ccccc}
s_{1}^{2} & s_{12} & s_{13} & \cdots & s_{1 p} \\
s_{21} & s_{2}^{2} & s_{23} & \cdots & s_{2 p} \\
s_{31} & s_{32} & s_{3}^{2} & \cdots & s_{3 p} \\
\vdots & \vdots & \vdots & \ddots & \vdots \\
s_{p 1} & s_{p 2} & s_{p 3} & \cdots & s_{p}^{2}
\end{array}\right)
$$
$$
s_{j}^{2}=(1 / n) \sum_{i=1}^{n}\left(x_{i j}-\bar{x}_{j}\right)^{2} \text { is the variance of the } j \text {-th variable }
$$
$$
\begin{aligned}
&s_{j k}=(1 / n) \sum_{i=1}^{n}\left(x_{i j}-\bar{x}_{j}\right)\left(x_{i k}-\bar{x}_{k}\right) \text { is the covariance between the } j \text {-th and } k \text {-th variables }
\end{aligned}
$$
$$
\bar{x}_{j}=(1 / n) \sum_{i=1}^{n} x_{i j} \text { is the mean of the } j \text {-th variable }
$$

We can calculate the covariance matrix such as

$$
\mathbf{S}=\frac{1}{n} \mathbf{X}_{c}^{\prime} \mathbf{X}_{c}
$$

Note that the centered matrix $\mathbf{X}_{c}$

$$
\mathbf{X}_{c}=\left(\begin{array}{cccc}
x_{11}-\bar{x}_{1} & x_{12}-\bar{x}_{2} & \cdots & x_{1 p}-\bar{x}_{p} \\
x_{21}-\bar{x}_{1} & x_{22}-\bar{x}_{2} & \cdots & x_{2 p}-\bar{x}_{p} \\
x_{31}-\bar{x}_{1} & x_{32}-\bar{x}_{2} & \cdots & x_{3 p}-\bar{x}_{p} \\
\vdots & \vdots & \ddots & \vdots \\
x_{n 1}-\bar{x}_{1} & x_{n 2}-\bar{x}_{2} & \cdots & x_{n p}-\bar{x}_{p}
\end{array}\right)
$$

How?

```{r gna3}
# More direct
n <- nrow(X)
m <- matrix(1, n, 1)%*%colMeans(X)
Xc <- X-m
Xc

# Or
#http://users.stat.umn.edu/~helwig/notes/datamat-Notes.pdf
C <- diag(n) - matrix(1/n, n, n)
XC <- C %*% X
XC

# Or 
Xcc <- scale(X, center=TRUE, scale=FALSE)

# Covariance Matrix
S <- t(Xc) %*% Xc / (n-1)
S

# Check it
cov(X)
```

## Correlation

$$
\mathbf{R}=\left(\begin{array}{ccccc}
1 & r_{12} & r_{13} & \cdots & r_{1 p} \\
r_{21} & 1 & r_{23} & \cdots & r_{2 p} \\
r_{31} & r_{32} & 1 & \cdots & r_{3 p} \\
\vdots & \vdots & \vdots & \ddots & \vdots \\
r_{p 1} & r_{p 2} & r_{p 3} & \cdots & 1
\end{array}\right)
$$

where

$$
r_{j k}=\frac{s_{j k}}{s_{j} s_{k}}=\frac{\sum_{i=1}^{n}\left(x_{i j}-\bar{x}_{j}\right)\left(x_{i k}-\bar{x}_{k}\right)}{\sqrt{\sum_{i=1}^{n}\left(x_{i j}-\bar{x}_{j}\right)^{2}} \sqrt{\sum_{i=1}^{n}\left(x_{i k}-\bar{x}_{k}\right)^{2}}}
$$
is the Pearson correlation coefficient between variables $\mathbf{X}_{j}$ and $\mathbf{X}_{k}$

We can calculate the correlation matrix such as

$$
\mathbf{R}=\frac{1}{n} \mathbf{X}_{s}^{\prime} \mathbf{X}_{s}
$$

where $\mathbf{X}_{s}=\mathbf{C X D}^{-1}$ with
  
- $\mathbf{C}=\mathbf{I}_{n}-n^{-1} \mathbf{1}_{n} \mathbf{1}_{n}^{\prime}$ denoting a centering matrix, 
- $\mathbf{D}=\operatorname{diag}\left(s_{1}, \ldots, s_{p}\right)$ denoting a diagonal scaling matrix.  

Note that the standardized matrix $\mathbf{X}_{s}$ has the form

$$
\mathbf{X}_{s}=\left(\begin{array}{cccc}
\left(x_{11}-\bar{x}_{1}\right) / s_{1} & \left(x_{12}-\bar{x}_{2}\right) / s_{2} & \cdots & \left(x_{1 p}-\bar{x}_{p}\right) / s_{p} \\
\left(x_{21}-\bar{x}_{1}\right) / s_{1} & \left(x_{22}-\bar{x}_{2}\right) / s_{2} & \cdots & \left(x_{2 p}-\bar{x}_{p}\right) / s_{p} \\
\left(x_{31}-\bar{x}_{1}\right) / s_{1} & \left(x_{32}-\bar{x}_{2}\right) / s_{2} & \cdots & \left(x_{3 p}-\bar{x}_{p}\right) / s_{p} \\
\vdots & \vdots & \ddots & \vdots \\
\left(x_{n 1}-\bar{x}_{1}\right) / s_{1} & \left(x_{n 2}-\bar{x}_{2}\right) / s_{2} & \cdots & \left(x_{n p}-\bar{x}_{p}\right) / s_{p}
\end{array}\right)
$$

How?

```{r gna4}
# More direct
n <- nrow(X)
sdx <- 1/matrix(1, n, 1)%*%apply(X, 2, sd)
m <- matrix(1, n, 1)%*%colMeans(X)
Xs <- (X-m)*sdx
Xs

# Or
#http://users.stat.umn.edu/~helwig/notes/datamat-Notes.pdf
C <- diag(n) - matrix(1/n, n, n)
D <- diag(apply(X, 2, sd))
XS <- C %*% X %*% solve(D)
XS

# Or 
Xss <- scale(X, center=TRUE, scale=TRUE)

# Covariance Matrix
R <- t(Xs) %*% Xs / (n-1)
R

# Check it
cor(X)
```

## Precision matrix

The inverse of this matrix, $\mathrm{S}_{\mathrm{XX}}^{-1}$, if it exists, is the inverse covariance matrix, also known as the concentration matrix or **precision matrix**.  
Let us consider a 2×2 covariance matrix

$$
\left[\begin{array}{cc}
\sigma^{2}(x) & \rho \sigma(x) \sigma(y) \\
\rho \sigma(x) \sigma(y) & \sigma^{2}(y)
\end{array}\right]
$$

Its inverse:

$$
\frac{1}{\sigma^{2}(x) \sigma^{2}(y)-\rho^{2} \sigma^{2}(x) \sigma^{2}(y)}\left[\begin{array}{cc}
\sigma^{2}(y) & -\rho \sigma(x) \sigma(y) \\
-\rho \sigma(x) \sigma(y) & \sigma^{2}(x)
\end{array}\right]
$$
  
If call the the precision matrix $D$, the correlation coefficient will be   
  
$$
-\frac{d_{i j}}{\sqrt{d_{i i}} \sqrt{d_{j j}}},
$$
Or,
  
$$
\frac{-\rho \sigma_{x} \sigma_{y}}{\sigma_{x}^{2} \sigma_{y}^{2}\left(1-e^{2}\right)} \times \sqrt{\sigma_{x}^{2}\left(1-\rho^{2}\right)} \sqrt{\sigma_{y}^{2}\left(1-\rho^{2}\right)}=-\rho
$$
That was for a 2x2 cov-matrix.  When we have more columns, the correlation coefficient reflects partial correlations. Here is an example:  
  
```{r gna5}
pm <- solve(S) # precision matrix
pm

# Partial correlation of 1,2
-pm[1,2]/(sqrt(pm[1,1])*sqrt(pm[2,2])) 

# Or
-cov2cor(solve(S))

# Or
ppcor::pcor(X)
```

## Semi-partial correlation

With partial correlation, we find the correlation between X and Y holding Z constant for both X and Y. Sometimes, however, we want to hold Z constant for just X or just Y. In that case, we compute a semipartial correlation. A partial correlation is computed between two residuals. A semipartial is computed between one residual and another raw or unresidualized variable. One interpretation of the semipartial is that it is the correlation between one variable and the residual of another, so that the influence of a third variable is only partialed from one of two variables (hence, semipartial). Another interpretation is that the semipartial shows the increment in correlation of one variable above and beyond another. This is seen most easily with the R2 formulation.

Partial
$$
r_{12.3}^{2}=\frac{R_{1.23}^{2}-R_{1.3}^{2}}{1-R_{1.3}^{2}}
$$
  
SemiPartial

$$
r_{1(2.3)}^{2}=R_{1.23}^{2}-R_{1.3}^{2}
$$



The difference between a slope coefficient, semi-partial and partial correlation can be seen by looking their definitions:  
  
**Partial:**    
  
$$
x_{12,3}=\frac{r_{12}-r_{13} r_{23}}{\sqrt{1-r_{12}^{2}} \sqrt{1-r_{23}^{2}}}
$$

**Regression:**

$$
x_{1}=b_{1}+b_{2} x_{2}+b_{2} X_{3}
$$
and  

$$
b_{2}=\frac{\sum x_{3}^{2} \sum x_{1} x_{2}-\sum x_{1} x_{3} \sum x_{2} x_{3}}{\sum x_{2}^{2} \sum x_{3}^{2}-\left(\sum x_{2} x_{3}\right)^{2}}
$$

With standardized variables:

$$
b_{2}=\frac{r_{12}-r_{13} r_{23}}{1-r_{23}^{2}}
$$
  
**Semi-partial (or "part") correlation:**
   
$$
r_{1(2.3)}=\frac{r_{1 2}-r_{1_{3}} r_{23}}{\sqrt{1-r_{23}^{2}}}
$$

see <http://faculty.cas.usf.edu/mbrannick/regression/Partial.html> for very nice Venn diagrams.

The difference is the square root in the denominator. The regression coefficient can exceed 1.0 in absolute value; the correlation cannot.  The difference between "beta" coefficients and semipartial is that semipartial normalizes the coefficient between -1 and +1.  The function `spcor` can calculate the pairwise semi-partial (part) correlations for each pair of variables given others. In addition, it gives us the p value as well as statistic for each pair of variables.


```{r gna6}
Xx <- X[,1]
Y <- X[,2]
Z <- X[,3]
ppcor::spcor(X)
lm(Xx~Y+Z)
```
  
For more information and matrix solutions to partial correlations, see <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4681537/pdf/nihms740182.pdf> [@Kim_2015].  

  

